---
title: "NLSY Parent Income and Wealth Imputation"
format:
  html:
    embed-resources: true
editor: visual
execute: 
  warning: false
fig-cap-location: top
toc: true
toc-expand: true
---

## Data Cleaning

```{r}
#Packages
librarian::shelf(dplyr, mice, car, corrplot, caret, tidyr)
source(here::here("NLSY/nlsy_lib.R"))

# Read data
# nlsydf = readRDS(paste0(here::here(), "/NLSY/NLSY-college-finance.rds"))
```

```{r}

basedf = nlsy_get_base_df()

# Making dummy or factor variables
test <- basedf |>
    mutate(
        mom_educ_hs = nlsy_encode_educ5(mom_res_ed, factorize=TRUE),
        dad_educ_hs = nlsy_encode_educ5(dad_res_ed, factorize=TRUE),
        race_eth = nlsy_recode_race_and_eth10(race, hisp),
        race_eth = factor(race_eth),
        race = factor(race),
        hisp = factor(hisp),
        par_dec = ifelse(par1_deceased == "Yes" | par2_deceased == "Yes", 1, 0) # Either parent deceased
    )

```

```{r}
# Looking for significance with non-missing income and wealth
non_missing <- test %>%
    filter(!is.na(pincome),
           !is.na(pnetworth))

summary(model <- lm(pincome ~ mom_age + hisp + race + mom_educ_hs + dad_educ_hs + has_retsav + owns_home + both_parents + par_dec, data = non_missing))
#Non-significant: par_dec - dropping for now
```

### Checking for multi-collinearity - correlation plot for numeric variables

```{r}

testing_cor <- test |>
    select(age_resp, mom_age, has_retsav, owns_home, both_parents) |>
    drop_na()

M <- cor(testing_cor)
cor(testing_cor)
corrplot(M, type = "upper")
```

### Checking for multi-collinearity - VIF and Successive addition of regressors

```{r}

summary(model <- lm(pincome ~ mom_age, data = test)) #Adj R2 = 0.057
summary(model <- lm(pincome ~ mom_age + hisp, data = test)) #Adj R2 = 0.09
car::vif(model)
summary(model <- lm(pincome ~ mom_age + hisp + race, data = test)) # #Adj R2 = 0.157
car::vif(model)

summary(model <- lm(pincome ~ mom_age + hisp + race + mom_educ_hs, data = test)) #Adj R2 = 0.1899
car::vif(model)

summary(model <- lm(pincome ~ mom_age + hisp + race + mom_educ_hs + dad_educ_hs, data = test)) #Adj R2 = 0.1616
car::vif(model)

summary(model <- lm(pincome ~ mom_age + hisp + race + mom_educ_hs + dad_educ_hs + has_retsav, data = test)) #Adj R2 = 0.199
car::vif(model)

summary(model <- lm(pincome ~ mom_age + hisp + race + mom_educ_hs + dad_educ_hs + has_retsav + owns_home, data = test)) #Adj R2 = 0.2107
car::vif(model)

summary(model <- lm(pincome ~ mom_age + hisp + race + mom_educ_hs + dad_educ_hs + has_retsav + owns_home + both_parents, data = test)) #Adj R2 = 0.2108
car::vif(model)
```

## Checking for missings in the predictor variables

```{r}

impute <- test |>
    select(
        id, race, pincome, pnetworth, mom_age, mom_educ_hs, dad_educ_hs, 
        hisp, race_eth, has_retsav, owns_home, both_parents
    )

# no variables are found to have near zero variance
t <- nearZeroVar(impute) 

# dad_educ_hs has 4598 missings
# TODO: dad_educ_hs has 3433 missings in data-cleaning
# and 771 missings in data-cleaning-refactor
sapply(impute, function(x) sum(is.na(x))) 

# Calculating percent of usable cases
m <- md.pairs(impute)

# looking for low proportions: where both target and predictor are missing on the same cases
# Note: For row target pincome and pnetworth respectively, predictors dad_educ_hs, has_retsav, 
# and owns_home have lowest proportions
round(m$mr/(m$mr + m$mm), 3) 

# I suspect multi-collinearity is causing singularity in the multiple imputation below, 
# so removed race and only kept interaction race_eth for now
impute <- impute %>%
    select(-race, -hisp) 
```

## Multiple Imputation

```{r}
# Paper that uses multiple imputation: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6049093/
# Multiple imputation in R reference: https://library.virginia.edu/data/articles/getting-started-with-multiple-imputation-in-r
init = mice(impute, maxit=5)
meth = init$method
predM = init$predictorMatrix

# Set distributions
num <- c("mom_age") # numerical variables
log <- c("has_retsav", "owns_home", "both_parents") # binary variables
poly <- c("mom_educ_hs") # ordered categorical variables
poly2 <- c("race", "race_eth")# Unordered categorical variable

# Assign Variables to be imputed
predM[,c("id")] <- 0 # not using as a predictor variable
predM[c("id"),] <- 0 # not using as a predictor variable
predM[c("race_eth"),] <- 1 # not using as a predictor variable
#predM[,c("hisp", "mom_age", "has_retsav", "owns_home", "mom_educ_hs", "both_parents", "race_eth")] <- 0 
#meth[c("hisp", "mom_age", "has_retsav", "owns_home", "mom_educ_hs", "both_parents", "race_eth")]= "" 
meth
predM
```

```{r}
# Income and Wealth imputation
set.seed(25)
imputed = mice(impute, method=meth, predictorMatrix=predM, m=5)

imputed <- complete(imputed, action = "repeated", include = FALSE)

final <- imputed %>%
    rename(id = id.1,
           race_eth = race_eth.1) %>%
    select(-starts_with("id."), -starts_with("race_eth"))
```

```{r}
# Join back with original income and wealth
og_values <- impute %>%
    select(id, pincome, pnetworth, race_eth, dad_educ_hs)

final <- merge(x=final,y=og_values,
                    by="id", all.x=TRUE)

income <- final %>%
    select(id, starts_with("pincome"))
wealth <- final %>%
    select(id, starts_with("pnetworth"))
dad_ed <- final %>%
    select(id, starts_with("dad_"))

pincome <- tidyr::gather(income, key = "imputation", value ="pincome", matches("pincome"), -id)
pnetworth <- tidyr::gather(wealth, key = "imputation", value ="pnetworth", matches("pnetworth"), -id)
dad_educ  <- tidyr::gather(dad_ed, key = "imputation", value ="dad_educ", matches("dad_educ"), -id)
dad_educ = mutate(dad_educ, dad_educ = factor(dad_educ, levels=nlsy_get_educ5_levels()))
```

## Distribution Plots for Parent Income (non-imputed and imputed)

```{r}
ggplot(pincome, aes(x=pincome, fill= imputation)) +
    geom_density(alpha = 0.5) +
    facet_wrap(~imputation, scales="free")
```

## Distribution Plots for Parent Wealth (non-imputed and imputed)

```{r}
ggplot(pnetworth, aes(x=pnetworth, fill= imputation)) +
    geom_density(alpha = 0.5) +
    facet_wrap(~imputation, scales="free")
```

## Distribution Plots for Dad Education (non-imputed and imputed)

```{r}

tmpdf = dad_educ |>
    filter(if_all(everything(), ~!is.na(.) )) |>
    mutate(imputation = if_else(imputation=="dad_educ_hs", "0", imputation),
           imputation = gsub("dad_educ_hs.", "", imputation))


tmpdf |>
    group_by(imputation) |>
    mutate(N=n()) |>
    group_by(dad_educ, imputation) |>
    summarise(N=max(N), share=n()/N) |>
    ggplot() +
    geom_col(
        aes(y=share, x=imputation, fill=dad_educ), 
        position = position_stack(reverse = TRUE)
        )


tmpdf |>
    left_join(select(test, id, race_eth), by="id") |>
    filter(!is.na(race_eth)) |>
    group_by(race_eth, imputation) |>
    mutate(N=n()) |>
    group_by(dad_educ, race_eth, imputation) |>
    summarise(N=max(N), share=n()/N) |>
    ggplot() +
    geom_col(
        aes(y=share, x=imputation, fill=dad_educ),
        position = position_stack(reverse = TRUE)
        ) +
    facet_wrap(~race_eth, nrow=2)

```

## Whisker Plots

```{r}
boxplot(pincome$pincome ~ pincome$imputation)
boxplot(pnetworth$pnetworth ~ pnetworth$imputation)
```

## Scatter Plots for both parent income and wealth (non-imputed is zero)

```{r}
pincome$imputation <- gsub("pincome.", "", pincome$imputation)
pincome$imputation <- ifelse(pincome$imputation == "pincome", 0, pincome$imputation)

pnetworth$imputation <- gsub("pnetworth.", "", pnetworth$imputation)
pnetworth$imputation <- ifelse(pnetworth$imputation == "pnetworth", 0, pnetworth$imputation)

merge <- merge(pincome, pnetworth, by = c("id", "imputation"))

ggplot(merge, aes(x=pincome, y=pnetworth, color = imputation)) +
    geom_point(alpha = 0.5) +
    facet_wrap(~imputation, scales="free")
```
