---
title: "NLSY Data Exploration"
output: html_document
date: '2023-10-07'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(Hmisc)
library(tidyverse)
library(scales)
library(ggridges)
library(mice)
library(urbnthemes)

set_urbn_defaults(style = "print")
source(here::here("NLSY/nlsy_lib.R"))
```


```{r}
# Base frame with demographic characteristics
basedf = nlsy_get_base_df()
```


# NLSY Data

## Sample Composition

The racial and ethnic composition of the sample is following:

```{r}
knitr::kable(table(basedf$race, basedf$hisp, useNA="ifany"))

basedf = basedf |> 
  nlsy_recode_race()
```


Because the sample sizes for people who identify as American Indian, Eskimo, or Aleut and Asian or Pacific Islander are small, we group them with others. We also drop those who miss race information.

```{r}

basedf = basedf |> 
  filter(!is.na(race) & !is.na(hisp) & race!='No information')

knitr::kable(table(basedf$race, basedf$hisp, useNA="ifany"))
#knitr::kable(table(basedf$race, basedf$bdate_y, useNA="ifany"))

```


## Wealth and Income


The number of observations with missing parental wealth is `r basedf |> filter(is.na(pnetworth)) |> summarize(n=n())`, those that are missing parental income is `r basedf |> filter(is.na(pincome)) |> summarize(n=n())` and those that are missing either one is `r basedf |> filter(is.na(pnetworth) | is.na(pincome)) |> summarize(n=n())`.

```{r fig.cap="Missing Values for Parents' Net Worth and Income"}
md.pattern(select(basedf, pincome, pnetworth))
```


```{r fig.cap="Distribution of Parents' Net Worth by Race (Thousands od Dollars)"}

basedf |>
  select(id, pnetworth, race, wt) |>
  mutate(wt=as.integer(wt/100)) |>
  uncount(wt) |>
  ggplot() +
  geom_boxplot(aes(y=pnetworth, color=race, fill=race)) +
  scale_y_continuous(labels=\(x) x/1000) +
  ylab("Net Worth ($1,000)") + remove_axis(axis='x') + remove_ticks()
```

```{r fig.cap="Distribution of Parents' Income by Race (Thousands od Dollars)"}

basedf |>
  select(id, pincome, race, wt) |>
  mutate(wt=as.integer(wt/100)) |>
  uncount(wt) |>
  ggplot() +
  geom_boxplot(aes(y=pincome, color=race, fill=race)) +
  scale_y_continuous(labels=\(x) x/1000) +
  ylab("Income ($1,000)") + remove_axis(axis='x') + remove_ticks()
```

```{r fig.cap="Parents' Net Worth and Gross Income by Race (Thousands of Dollars)"}

basedf |>
  ggplot(aes(x=pnetworth, y=pincome, color=race)) +
    geom_point(size=1) + 
    scale_y_continuous(labels=\(x) x/1000) +
    scale_x_continuous(labels=\(x) x/1000) +
    xlab("Net Worth ($1,000)") + ylab('Income ($1,000)')

```

```{r fig.cap="Distribution of Parents' Net Worth by Race (Thousands of Dollars)"}

basedf |>
  select(id, pnetworth, pincome, race, wt) |>
  mutate(wt=as.integer(wt/100)) |>
  uncount(wt) |>
  ggplot() +
  geom_freqpoly(aes(x=pnetworth, y=after_stat(density), color=race)) +
  scale_x_continuous(labels=\(x) x/1000) +
  xlab("Net worth ($1,000)") + ylab("Density")


```

# College Education

## Ever Attended College

```{r}
# Frame with school enrollment
schdf = nlsy_get_col_stat_annual_df()

# Frame with ever attended college
evercoldf = schdf |>
  group_by(id) |>
  summarise(evercol = max(colenr))

# Sample of 
table(evercoldf$evercol)
# 0    1 
# 3606 5378 

# Merge with basedf
evercoldf = left_join(
  evercoldf,
  basedf,
  by='id'
)

```

```{r fig.cap="Share of Ever Attended College by Parents' Net-Worth Quartile and Race"}


evercoldf |>
  filter(!is.na(pnetworth)) |>
  mutate(
    pnworth_cat = cut(
      pnetworth, 
      breaks=Hmisc::wtd.quantile(pnetworth, wt, probs=seq(0,1,.2)),
      labels=c('1st','2nd','3rd','4th','5th'),
      include.lowest = TRUE)
  ) |>
  group_by(race, pnworth_cat) |>
  mutate(
    `College Attendance`=weighted.mean(evercol, wt),
    se = sqrt(`College Attendance`*(1-`College Attendance`)/n()),
    ci_l = `College Attendance`-1.96*se,
    ci_h = `College Attendance`+1.96*se) |>
  ggplot() +
    geom_col(aes(x=pnworth_cat,y=`College Attendance`,fill=race), position="dodge") +
    geom_errorbar(aes(x=pnworth_cat, ymin=ci_l, ymax=ci_h, group=race), color='red', width = .2, position=position_dodge(.7)) +
    xlab("Net Worth Quintile")
```

```{r fig.cap="Share of Ever Attended College by Parents' Income Quartile and Race"}



evercoldf |>
  filter(!is.na(pincome)) |>
  mutate(
    pincome_cat = cut(
      pincome, 
      breaks=Hmisc::wtd.quantile(pincome, wt, probs=seq(0,1,.2)), 
      labels=c('1st','2nd','3rd','4th','5th'),
      include.lowest = TRUE)
  ) |>
  group_by(race, pincome_cat) |>
  mutate(`College Attendance`=weighted.mean(evercol, wt),
    se = sqrt(`College Attendance`*(1-`College Attendance`)/n()),
    ci_l = `College Attendance`-1.96*se,
    ci_h = `College Attendance`+1.96*se) |>
  ggplot() +
    geom_col(aes(x=pincome_cat,y=`College Attendance`,fill=race), position="dodge") +
    geom_errorbar(aes(x=pincome_cat, ymin=ci_l, ymax=ci_h, group=race), color='red', width = .2, position=position_dodge(.7)) +
    xlab("Income Quintile")
```

## College Enrollment History


```{r}
# Frame with college enrollment
colstdf = nlsy_get_col_stat_fall_df()

# Frame with highest grade completed
hgcdf = nlsy_get_highest_grade_completed_df()

colstdf = left_join(
  colstdf,
  hgcdf,
  by=c('id', 'year')
)

# Do not use this!
#
# This way of constructing the year of graduation from high-school is inconsistent with
# the CV_HGC_EVER_ variables. Using it results in people who graduated from high school 
# and completed first year of college in the same year.
#
# The year in which graduated from high school
# Expressed as number of months since 1/1/1980.
# If diploma was obtained in August or later, increment year
# hscompdf = nlsydf |>
#   select(
#     id  = "PUBID_1997",
#     hs  = "CVC_HS_DIPLOMA_XRND",
#     ged = "CVC_GED_XRND"
#   ) |>
#   mutate(
#     hs_comp_date = case_when(
#       !is.na(hs)  ~ hs,
#       !is.na(ged) ~ ged,
#       TRUE ~ NA
#     )
#   ) |>
#   mutate(
#     hs_comp_year = 1980 + hs_comp_date %/% 12,
#     hs_comp_month = hs_comp_date %% 12,
#     hs_comp_year = ifelse(hs_comp_month >= 8, hs_comp_year+1, hs_comp_year)
#     ) |>
#   select(id, hs_comp_year) |>
#   filter(!is.na(hs_comp_year))
# 
# colstdf = left_join(
#   colstdf,
#   hscompdf,
#   by=c('id')
# )

colstdf = colstdf |>
  filter(year>=hs_comp_year) |>
  mutate(
    cumenr = cumsum(enrolled),
    col1 = as.integer(enrolled==TRUE & cumenr==1),
    year_since_hs = year-hs_comp_year
    )


```


This chart shows the number of college years completed in every year of the survey since the graduation from high school. Each row represents an NLSY participant who graduated from high school and the x-axis represents the number of years since the graduation. Cells are color-coded according to the number of years of college completed.

```{r fig.width=7, fig.height=10}

plotdf = colstdf |> 
  ungroup() |>
  select(id, year_since_hs, hcyc) |>
  pivot_wider(id_cols=id, names_from=year_since_hs, values_from = hcyc, names_prefix="y") |>
  arrange(y0,y1,y2,y3,y4,y5,y6,y7,y8,y9,y10,y11,y12,y13,y14,y15,y16,y17,y18,y19,y20,y21,y22) |>
  mutate(id2=row_number()) |>
  pivot_longer(
    starts_with('y'),
    names_prefix='y',
    values_to = 'hcyc',
    names_to='year_since_hs') |>
  mutate(
    year_since_hs=as.integer(year_since_hs)
    ) 

plotdf |>
  ggplot() + 
    geom_tile(aes(x=year_since_hs, y=id2, fill=hcyc)) +
    theme() + 
    scale_fill_gradientn() +
    theme(legend.position = "right",
          legend.direction = "vertical",
          axis.line.x = element_blank(),
          axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank(),
          panel.grid.major.y = element_blank()) +
    remove_ticks() +
    xlab("Years Since High School")

```


# Student loans

```{r}
# Student loans
sloandf = nlsy_get_student_loans_df()

sloandf = left_join(
  basedf |> select(id, bdate_y, sex, race, hisp, wt),
  sloandf,
  by='id'
)
```

```{r}
  
plotdf = sloandf |>
  filter(hasdebt==1) |>
  group_by(year, race) |>
  summarise(
    debtavg=weighted.mean(debt, wt),
    debtmed=median(debt)
    )
  
plotdf |>
  ggplot() +
    geom_line(aes(x=year, y=debtavg, color=race, group=race))

plotdf |>
  ggplot() +
  geom_line(aes(x=year, y=debtmed, color=race, group=race))

```

# Housing

## Living with Parents

The NLSY survey defines independence fairly broadly. To better understand the degree of dependence on parents, we construct a variable that is true if any of the following lives in the same household as the participant:

* Mother,
* Father,
* Adoptive mother,                          
* Adoptive father,
* Step-mother,
* Step-father,                              
* Foster mother,
* Foster father

From https://www.nlsinfo.org/content/cohorts/nlsy97/topical-guide/income/income:

>To be considered independent, a respondent had at least one of these characteristics: was age 18 or older, had a child, was enrolled in a 4-year college, was or had been married or was in a marriage-like relationship (defined in rounds 1-8 as a sexual relationship where partners of the opposite sex live together) at the time of the survey, was no longer enrolled in school, or was not living with any parents or parent-figures. Beginning in round 8, all respondents were classified as independent.

```{r}

withpardf = nlsy_get_famrel_df() |>
  mutate(with_parents = (value %in% c(
      "Mother",
      "Father",
      "Adoptive mother",                          
      "Adoptive father",
      "Step-mother",
      "Step-father",                              
      "Foster mother",
      "Foster father"
    ))
  ) |>  
  group_by(id, year) |>
  dplyr::summarize(with_parents = min(1,sum(with_parents, na.rm = TRUE)))

withpardf = right_join(
  withpardf,
  select(basedf, id, bdate_y, race, wt),
  by='id'
) |>
  mutate(age=year-bdate_y)

plotdf = bind_rows(
  withpardf |>
    group_by(age) |>
    dplyr::summarize(with_parents=weighted.mean(with_parents, wt)) |>
    mutate(race = 'All'),
  withpardf |>
    mutate(race=as.character(race)) |>
    group_by(age,race) |>
    dplyr::summarize(with_parents=weighted.mean(with_parents, wt))
) 

ggplot(plotdf) +
  geom_line(
    mapping=aes(x=age,y=with_parents,color=race),
    size=1
    ) +
  scale_y_continuous(labels=\(x) x*100) +
  ylab("Living with Parents (%)") + xlab('Age')

write_csv(plotdf, 'live_w_parents.csv')

```
