---
title: "NLSY Data Exploration"
output: html_document
date: '2023-10-07'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(Hmisc)
library(tidyverse)
library(scales)
library(ggridges)
library(mice)
library(urbnthemes)

set_urbn_defaults(style = "print")
set.seed(8984)
```

```{r}
# Read data
nlsydf = readRDS(paste0(here::here(), "/NLSY/NLSY-college-finance.rds"))

```


```{r}
# Base frame with demographic characteristics
basedf = nlsydf |>
  select(
    id =        "PUBID_1997",
    age0 =      "CV_AGE_12/31/96_1997",
    sex =       "KEY_SEX_1997",
    bdate_m =   "KEY_BDATE_M_1997",
    bdate_y =   "KEY_BDATE_Y_1997",
    hisp =      "KEY_ETHNICITY_1997",
    race =      "KEY_RACE_1997",
    pincome =   "CV_INCOME_GROSS_YR_1997",
    pnetworth = "CV_HH_NET_WORTH_P_1997",
    has_retsav= "P5-130_1997",
    retsav1 =   "P5-131_1997",
    retsav2 =   "P5-132_1997",
    wt
  )

basedf = basedf |> 
  mutate(
    race = droplevels(race),
    hisp = fct_recode(
      hisp,
      'Hispanic'     = 'Yes',
      'Non-Hispanic' = 'No'
    ),
    has_retsav = case_when(
      has_retsav=='YES' ~ 1,
      has_retsav=='NO'  ~ 0,
      TRUE ~ NA
    )
  )

basedf = basedf |>
  mutate(
    retsav3 = case_when(
    retsav2 == "A.  $1               -       $5,000" ~ runif(n(), 1, 5000),
    retsav2 == "B.   $5,001      -     $10,000"      ~ runif(n(), 5001, 10000),
    retsav2 == "C.   $10,001    -     $25,000"       ~ runif(n(), 10001, 25000),
    retsav2 == "D.   $25,001    -     $50,000"       ~ runif(n(), 25001, 50000),
    retsav2 == "E.    $50,001   -    $100,000"       ~ runif(n(), 50001, 100000),
    retsav2 == "F.    $100,001       $250,000"       ~ runif(n(), 100001, 250000),
    retsav2 == "G.    More than $250,000"            ~ runif(n(), 250001, 500000),
    TRUE ~ NA
    )
  )

basedf = basedf |>
  mutate(
    retsav = case_when(
      !is.na(retsav1) ~ retsav1,
      !is.na(retsav3) ~ retsav3,
      TRUE ~ NA
    )
  )

# Check that we have retsav for all people who have retirement savings and
# whose networth is not missing
stopifnot(
  dim(
    filter(
      basedf, 
      !is.na(pnetworth) & !is.na(has_retsav) & has_retsav & is.na(retsav)
      )
  )[1] == 0
)
```


# NLSY Data

## Sample Composition

The racial and ethnic composition of the sample is following:

```{r}
knitr::kable(table(basedf$race, basedf$hisp, useNA="ifany"))

basedf = basedf |> 
  mutate(
    race = fct_recode(
      race,
      'Other'  = "Something else? (SPECIFY)",
      'Black'  = "Black or African American",
      'Other'  = "Asian or Pacific Islander",
      'Other' = "American Indian, Eskimo, or Aleut"
    )
  )
```


Because the sample sizes for people who identify as American Indian, Eskimo, or Aleut and Asian or Pacific Islander are small, we group them with others. We also drop those who miss race information.

```{r}

basedf = basedf |> 
  filter(!is.na(race) & !is.na(hisp) & race!='No information')

knitr::kable(table(basedf$race, basedf$hisp, useNA="ifany"))
#knitr::kable(table(basedf$race, basedf$bdate_y, useNA="ifany"))

```


## Wealth and Income


The number of observations with missing parental wealth is `r basedf |> filter(is.na(pnetworth)) |> summarize(n=n())`, those that are missing parental income is `r basedf |> filter(is.na(pincome)) |> summarize(n=n())` and those that are missing either one is `r basedf |> filter(is.na(pnetworth) | is.na(pincome)) |> summarize(n=n())`.

```{r fig.cap="Missing Values for Parents' Net Worth and Income"}
md.pattern(select(basedf, pincome, pnetworth))
```


```{r fig.cap="Distributiuon of Parents' Net Worth by Race (Thousands od Dollars)"}

basedf |>
  select(id, pnetworth, race, wt) |>
  mutate(wt=as.integer(wt/100)) |>
  uncount(wt) |>
  ggplot() +
  geom_boxplot(aes(y=pnetworth, color=race, fill=race)) +
  scale_y_continuous(labels=\(x) x/1000) +
  ylab("Net Worth ($1,000)") + remove_axis(axis='x') + remove_ticks()
```

```{r fig.cap="Distributiuon of Parents' Income by Race (Thousands od Dollars)"}

basedf |>
  select(id, pincome, race, wt) |>
  mutate(wt=as.integer(wt/100)) |>
  uncount(wt) |>
  ggplot() +
  geom_boxplot(aes(y=pincome, color=race, fill=race)) +
  scale_y_continuous(labels=\(x) x/1000) +
  ylab("Income ($1,000)") + remove_axis(axis='x') + remove_ticks()
```

```{r fig.cap="Parents' Net Worth and Gross Income by Race (Thousands of Dollars)"}

basedf |>
  ggplot(aes(x=pnetworth, y=pincome, color=race)) +
    geom_point(size=1) + 
    scale_y_continuous(labels=\(x) x/1000) +
    scale_x_continuous(labels=\(x) x/1000) +
    xlab("Net Worth ($1,000)") + ylab('Income ($1,000)')

```

```{r fig.cap="Distribution of Parents' Net Worth by Race (Thousands of Dollars)"}

basedf |>
  select(id, pnetworth, pincome, race, wt) |>
  mutate(wt=as.integer(wt/100)) |>
  uncount(wt) |>
  ggplot() +
  geom_freqpoly(aes(x=pnetworth, y=after_stat(density), color=race)) +
  scale_x_continuous(labels=\(x) x/1000) +
  xlab("Net worth ($1,000)") + ylab("Density")


```

## College Attendance

```{r}
# Frame with school enrollment
schdf = nlsydf |>
  select(
    id =        "PUBID_1997",
    starts_with('CV_ENROLLSTAT')
  ) |>
  rename_with(~gsub('_EDT_', '_', .x)) |>
  pivot_longer(
    starts_with('CV_ENROLLSTAT'), 
    names_to='year', 
    names_prefix='CV_ENROLLSTAT_') |>
  mutate(year=as.integer(year)) |>
  mutate(colenr = case_when(
    (value=="Enrolled in a 2-year college" |
     value=="Enrolled in a 4-year college") ~ 1,
    TRUE ~ 0
    ) 
  )

col1styrdf = schdf |>
  group_by(id) |>
  filter(colenr==1) |>
  filter(row_number()==1) |>
  rename(first_col_year=year)

# Frame with ever attended college
evercoldf = schdf |>
  group_by(id) |>
  summarise(evercol = max(colenr))

# Sample of 
table(evercoldf$evercol)
# 0    1 
# 3606 5378 

evercoldf = left_join(
  evercoldf,
  basedf,
  by='id'
)



```

```{r fig.cap="Share of Ever Attended College by Parents' Net-Worth Quartile and Race"}



evercoldf |>
  filter(!is.na(pnetworth)) |>
  mutate(
    pnworth_cat = cut(
      pnetworth, 
      breaks=Hmisc::wtd.quantile(pnetworth, wt, probs=seq(0,1,.2)),
      labels=c('1st','2nd','3rd','4th','5th'),
      include.lowest = TRUE)
  ) |>
  group_by(race, pnworth_cat) |>
  mutate(
    `College Attendance`=weighted.mean(evercol, wt),
    se = sqrt(`College Attendance`*(1-`College Attendance`)/n()),
    ci_l = `College Attendance`-1.96*se,
    ci_h = `College Attendance`+1.96*se) |>
  ggplot() +
    geom_col(aes(x=pnworth_cat,y=`College Attendance`,fill=race), position="dodge") +
    geom_errorbar(aes(x=pnworth_cat, ymin=ci_l, ymax=ci_h, group=race), color='red', width = .2, position=position_dodge(.7)) +
    xlab("Net Worth Quintile")
```

```{r fig.cap="Share of Ever Attended College by Parents' Income Quartile and Race"}



evercoldf |>
  filter(!is.na(pincome)) |>
  mutate(
    pincome_cat = cut(
      pincome, 
      breaks=Hmisc::wtd.quantile(pincome, wt, probs=seq(0,1,.2)), 
      labels=c('1st','2nd','3rd','4th','5th'),
      include.lowest = TRUE)
  ) |>
  group_by(race, pincome_cat) |>
  mutate(`College Attendance`=weighted.mean(evercol, wt),
    se = sqrt(`College Attendance`*(1-`College Attendance`)/n()),
    ci_l = `College Attendance`-1.96*se,
    ci_h = `College Attendance`+1.96*se) |>
  ggplot() +
    geom_col(aes(x=pincome_cat,y=`College Attendance`,fill=race), position="dodge") +
    geom_errorbar(aes(x=pincome_cat, ymin=ci_l, ymax=ci_h, group=race), color='red', width = .2, position=position_dodge(.7)) +
    xlab("Income Quintile")
```

# Student loans

```{r}
# Student loans
sloandf = nlsydf |>
  select(
    id =        "PUBID_1997",
    starts_with('YSCH-25700')
  ) |>
  left_join(evercoldf, by='id') |>
  filter(evercol==1) |>
  select(-"YSCH-25700.01_1997", -evercol) |>
  pivot_longer(
    starts_with('YSCH-25700'), 
    names_to=c('college', 'term', 'year'),
    names_pattern='YSCH-25700\\.(\\d\\d)\\.(\\d\\d)\\_(\\d\\d\\d\\d)'
    ) |>
  group_by(id, year) |>
  summarise(debt=sum(value, na.rm=TRUE)) |>
  mutate(hasdebt=ifelse(debt>0, 1, 0))
  

sloandf = left_join(
  basedf |> select(id, bdate_y, sex, race, hisp, wt),
  sloandf,
  by='id'
)
```

```{r}
  
plotdf = sloandf |>
  filter(hasdebt==1) |>
  group_by(year, race) |>
  summarise(
    debtavg=weighted.mean(debt, wt),
    debtmed=median(debt)
    )
  
plotdf |>
  ggplot() +
    geom_line(aes(x=year, y=debtavg, color=race, group=race))

plotdf |>
  ggplot() +
  geom_line(aes(x=year, y=debtmed, color=race, group=race))

```

# Housing

## Living with Parents

The NLSY survey defines independence fairly broadly. To better understand the degree of dependence on parents, we construct a variable that is true if any of the following lives in the same household as the participant:

* Mother,
* Father,
* Adoptive mother,                          
* Adoptive father,
* Step-mother,
* Step-father,                              
* Foster mother,
* Foster father

From https://www.nlsinfo.org/content/cohorts/nlsy97/topical-guide/income/income:

>To be considered independent, a respondent had at least one of these characteristics: was age 18 or older, had a child, was enrolled in a 4-year college, was or had been married or was in a marriage-like relationship (defined in rounds 1-8 as a sexual relationship where partners of the opposite sex live together) at the time of the survey, was no longer enrolled in school, or was not living with any parents or parent-figures. Beginning in round 8, all respondents were classified as independent.

```{r}

withpardf = nlsydf |>
  select(
    id =        "PUBID_1997",
    matches('^HHI2*_RELY.*')
  ) |>
  rename_with(~gsub('HHI2_', 'HHI_',.x), starts_with('HHI2_')) |>
  pivot_longer(
    starts_with('HHI_'), 
    names_to=c('memberID', 'year'),
    names_pattern='HHI_RELY\\.(\\d\\d)_(\\d\\d\\d\\d)',
    names_transform = list(year=as.integer)
    ) |>
  mutate(with_parents = (value %in% c(
      "Mother",
      "Father",
      "Adoptive mother",                          
      "Adoptive father",
      "Step-mother",
      "Step-father",                              
      "Foster mother",
      "Foster father"
    ))
  ) |>  
  group_by(id, year) |>
  dplyr::summarize(with_parents = min(1,sum(with_parents, na.rm = TRUE)))

withpardf = right_join(
  withpardf,
  select(basedf, id, bdate_y, race, wt),
  by='id'
) |>
  mutate(age=year-bdate_y)

plotdf = bind_rows(
  withpardf |>
    group_by(age) |>
    dplyr::summarize(with_parents=weighted.mean(with_parents, wt)) |>
    mutate(race = 'All'),
  withpardf |>
    mutate(race=as.character(race)) |>
    group_by(age,race) |>
    dplyr::summarize(with_parents=weighted.mean(with_parents, wt))
) 

ggplot(plotdf) +
  geom_line(
    mapping=aes(x=age,y=with_parents,color=race),
    size=1
    ) +
  scale_y_continuous(labels=\(x) x*100) +
  ylab("Living with Parents (%)") + xlab('Age')

write_csv(plotdf, 'live_w_parents.csv')

```
